---
title: "Computer Lab 6 Computational Statistics"
author: "Phillip H??lscher"
date: "1 3 2019"
output: 
  pdf_document:
    toc: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,echo=FALSE}
# library used in this lab
library(ggplot2) # ex 2.1 - time series plot
library(gridExtra) 
```


\newpage

# Question 1: Genetic algorithm
```{r, echo=FALSE}
# clean the environment
rm(list=ls())
```


In this assignment, you will try to perform one-dimensional maximization with the help of a genetic algorithm.

## 1. Define the function

$$  f(x):= \frac{x^2}{e^x}- 2exp(\frac{-(9sin \  x)}{x^2+x+1} )$$

```{r}
# define the function
func = function(x){
  return((x^2/exp(x)) - 2 * exp(-(9 * sin(x))/ (x^2 +x +1)))
}
```

## 2. Define the function `crossover()`: 
for two scalars *x* and *y* it returns their "kid as (x+y)/2.

```{r}
# crossover function
crossover = function(x,y){
  kid = (x+y)/2
  return(kid)
}
```

## 3. Define the function 'mutate()' 
that for a scalar x returns the result of the integer division
x2 mod 30. (Operation mod is denoted in R as %%).

```{r}
# mutate function
mutate = function(x){
  return(x^2 %%30)
}
```

## 4. Write a function that depends on the parameters `maxiter` and `mutprob` and:

* (a) Plots function f in the range from 0 to 30. Do you see any maximum value?

* (b) Defines an initial population for the genetic algorithm as $X = (0, 5, 10, 15, . . . , 30)$.

* (c) Computes vector `Values` that contains the function values for each population point. 

* (d) Performs `maxiter` iterations where at each iteration

    + i. Two indexes are randomly sampled from the current population, they are further used as parents (use *sample()*).

    + ii. One index with the smallest objective function is selected from the current population, the point is referred to as victim (use *order()*).

    + iii. Parents are used to produce a new kid by crossover. Mutate this kid with probability *mutprob* (use *crossover()*, *mutate()*).

    + iv. The victim is replaced by the kid in the population and the vector *Values* is updated.

    + v. The current maximal value of the objective function is saved.


* (e) Add the final observations to the current plot in another colour.

```{r}
# ex4
# function that depends on the parameters `maxiter` and `mutprob`
set.seed(1234567)
func4 = function(maxiter, mutprob){
#test
# maxiter = 10
# mutprob = 0.1
  
# ------ a) #### -  Plots function f in the range
# create data for ggplot
range = c(0:30)
func_range = func(0:30)
# create plot data
plot_data1a = data.frame("func_range" = func_range,
                        "range" = range)
plot1a = ggplot(data = plot_data1a, aes(x = range, y = func_range)) + 
  geom_line() +
  #geom_point()+
  ggtitle("f in the range from 0 to 30") + xlab("range") + ylab("f")
plot1a
# ------ b) #### - initial population
population = seq(0,30, 5)

# ------ c) #### - Computes vector Values
Values = func(population)
max_objfunc = c()
# ------ d) #### - Performs maxiter
for (i in 1:maxiter) {
  # ------ d) ##### ------ i) ####
  index1 = sample(population,1)
  index2 = sample(population,1)
  parent = c(index1, index2)
# ------ d) ##### ------ ii) ####
  smalles_obfunc = order(Values)[1]
  victim = population[smalles_obfunc]
# ------ d) ##### ------ iii) ####
  new_kid = crossover(parent[1],parent[2])
  mutate_probability = runif(1,min = 0, max = 1)
  if(mutate_probability <= mutprob){ # >= OR <=
    new_kid = mutate(new_kid)
  } 
# ------ d) ##### ------ iv) ####
  replace_index = which(population == victim)
  population[replace_index] = new_kid
  Values = func(population)
# ------ d) ##### ------ v) ####
  max_objfunc[i] = max(Values)
}
# ------ e) #### - Add the final observations to the current plot in another colour
  #plot data
  plot_data1b = data.frame("Values" = Values,
                           "Population" = population)
  plot_data1b = plot_data1b[order(plot_data1b$Population),]
  col = c("Final observations" = "red")
  plot1b = plot1a + geom_point(data = plot_data1b, aes(x = Population, y = Values, colour = "Final observations"))
  plot1b
  return(plot1b)
}

```

```{r, echo=FALSE, eval=FALSE}

#### TEST -----------
#i. Two indexes are randomly sampled from the current population, they are further used as parents (use sample()).

sample(population, 1)

#ii. One index with the smallest objective function is selected from the current population, the point is referred to as victim (use order()).

order(parent)

# iii. Parents are used to produce a new kid by crossover. Mutate this kid with probability mutprob (use crossover(), mutate()).

new_kid = crossover(parent[1],parent[2])
probability = runif(1,0,1)
mutate(new_kid)

# iv. The victim is replaced by the kid in the population and the vector Values is updated.
which(population == 25)

# v. The current maximal value of the objective function is saved.
max_objfunc = max(Values)

```

5. Run your code with different combinations of **maxiter**= 10, 100 and **mutprob**= 0.1, 0.5, 0.9. Observe the initial population and final population. Conclusions?

```{r,echo=FALSE}
maxite = c(10,100)
mutpro = c(0.1,0.5,0.9)
# # amount of objects
# #amount_of_objects = length(maxite) * length(mutpro)
# all_resutls = list()
# counter = 1
# for (i in maxite) {
#   for (j in mutpro) {
#     # print(maxite[i])
#     # print(mutpro[j])
#     # print(counter)
#     all_resutls[counter] = func4(i,j)
#     counter = counter +1
#   }
# }
plot1 = func4(10,0.1) + ggtitle("maxiter = 10 & mutpro = 0.1")
plot2 = func4(10,0.5) + ggtitle("maxiter = 10 & mutpro = 0.5")
plot3 = func4(10,0.9) + ggtitle("maxiter = 10 & mutpro = 0.9")
plot4 = func4(100,0.1) + ggtitle("maxiter = 100 & mutpro = 0.1")
plot5 = func4(100,0.5) + ggtitle("maxiter = 100 & mutpro = 0.5")
plot6 = func4(100,0.9) + ggtitle("maxiter = 100 & mutpro = 0.9")

final_plot = grid.arrange(plot1,plot2,plot3,plot4,plot5,plot6, ncol=2)
final_plot
```





\newpage
#Question 2: EM algorithm

```{r, echo=FALSE}
# clean the environment
rm(list=ls())
# load the data
data2 = read.csv("physical1.csv")
```

The data file *physical.csv* describes a behavior of two related physical processes $Y = Y (X) and Z = Z(X)$.


## 1. Make a time series plot 
describing dependence of $Z$ and $Y$ versus $X$. Does it seem that two processes are related to each other? What can you say about the variation of the response values with respect to $X$?

```{r, echo=FALSE}
# Z & Y versus X
col = c("Y" = "#FFC312", "Z" = "#0652DD")

# ------- separate plots - together grid.arrange
# # create Y on X
# plot1_Y = ggplot(data = data2, aes(x = X)) +
#   geom_line(aes(y = Y, colour = "Y")) + 
#   ggtitle("Z & Y versus X") + ylab("Z & Y")
# 
# # create Z on X
# plot1_Z = ggplot(data = data2, aes(x = X)) +
#   geom_line(aes(y = Z, colour = "Z")) + 
#   ggtitle("Z & Y versus X") + ylab("Z & Y")
# 
# grid.arrange(plot1_Y,plot1_Z,nrow=2)

# create Y on X
plot1_YZ = ggplot(data = data2, aes(x = X)) +
  geom_line(aes(y = Y, colour = "Y")) + 
  geom_line(aes(y = Z, colour = "Z")) + 
  ggtitle("Z & Y versus X") + ylab("Z & Y")
plot1_YZ

```

It does not seem the two processes are related to each other.
In the beginning, the Z value does sleep off more than double of Y. 
Also, the movements of the processes rarely lie on top of each other. 
We can also recognize the Z values are incomplete in some parts, the curve has gaps. 

- variation of the response values with respect to X?


## 2. Note that there are some missing values of Z in the data 
which implies problems in estimating models by maximum likelihood. Use the following model

$$ Y_i \sim \rm exp(X_i/\lambda), \\  Z_i \sim \rm exp(X_i/2\lambda)$$

where $\lambda$ is some unknown parameter.
*The goal is to derive an EM algorithm that estimates $\lambda$.*


```{r}

```



\newpage

# Appendix
```{r ref.label=knitr::all_labels(), echo = T, eval = F}
```
