---
title: "Computer Lab 6 Computational Statistics"
author: "Phillip Hölscher (phiho267) & Zijie Feng (zijfe244)"
date: "1 3 2019"
output: 
  pdf_document:
    toc: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.height = "300px")
```

```{r,echo=FALSE}
# library used in this lab
library(ggplot2) # ex 2.1 - time series plot
library(gridExtra) 
```


\newpage

# Question 1: Genetic algorithm
```{r, echo=FALSE}
# clean the environment
rm(list=ls())
```


In this assignment, you will try to perform one-dimensional maximization with the help of a genetic algorithm.

## 1. Define the function

$$  f(x):= \frac{x^2}{e^x}- 2\exp(\frac{-9\sin x}{x^2+x+1} )$$

```{r}
# define the function
func = function(x){
  return((x^2/exp(x)) - 2 * exp(-(9 * sin(x))/ (x^2 +x +1)))
}
```

## 2. Define the function `crossover()`
for two scalars *x* and *y* it returns their "kid as $(x+y)/2$.

```{r}
# crossover function
crossover = function(x,y){
  kid = (x+y)/2
  return(kid)
}
```

## 3. Define the function 'mutate()' 
that for a scalar x returns the result of the integer division
x2 mod 30. (Operation mod is denoted in R as %%).

```{r}
# mutate function
mutate = function(x){
  return(x^2 %%30)
}
```

## 4. Write a function that depends on the parameters `maxiter` and `mutprob` and:

* (a) Plots function f in the range from 0 to 30. Do you see any maximum value?

* (b) Defines an initial population for the genetic algorithm as $X = (0, 5, 10, 15, . . . , 30)$.

* (c) Computes vector `Values` that contains the function values for each population point. 

* (d) Performs `maxiter` iterations where at each iteration

    + i. Two indexes are randomly sampled from the current population, they are further used as parents (use *sample()*).

    + ii. One index with the smallest objective function is selected from the current population, the point is referred to as victim (use *order()*).

    + iii. Parents are used to produce a new kid by crossover. Mutate this kid with probability *mutprob* (use *crossover()*, *mutate()*).

    + iv. The victim is replaced by the kid in the population and the vector *Values* is updated.

    + v. The current maximal value of the objective function is saved.


* (e) Add the final observations to the current plot in another colour.

```{r}
#1.4a#############################################################
dataa <- data.frame(x=0:30,f=func(0:30))
plot1.4=ggplot(dataa,aes(x=x,y=f,color="Red"))+
  geom_line()+
  geom_point()   # max x=1
plot1.4
```

```{r}
#1.4b#############################################################
# initial population
X <- seq(0,30,5)

#1.4c#############################################################
# the function values for each population points
Values <- func(X)
```

```{r}
#1.4d#############################################################
set.seed(1234567)
func4 <- function(pars, animation = F){
  maxiter = pars$maxiter
  mutprob = pars$mutprob
  name = pars$name
  tX <- X
  for(i in 1:maxiter) {
    samples <- sample(tX, 2, replace = F)
    id <- which.min(func(tX))
    kid <- crossover(samples[1],samples[2])
    if(runif(1)>mutprob){
      kid <- mutate(kid)
    }
    tX[id] <- kid
    tX <- sort(tX)
    if(animation){
      plot(tX,func(tX),type = "b",xlim=c(0,30), ylim = c(-3,0.25),col="Blue")
      lines(x=seq(0,30),y=f(seq(0,30)))
      Sys.sleep(0.2)
    }
  }
  dt <- data.frame(X=tX,f=func(tX))
  pl = ggplot(dt,aes(x=X,y=f,color="Blue"))+
    geom_point()+
    geom_line()+
    geom_line(data=dataa,aes(x=x,y=f,color="Red"))+
    ylim(-3,0.25)+
    xlim(0,30)+
    ggtitle(name)+
    scale_color_discrete(labels=c("GA","Func"))
  return(pl)
}
```

5. Run your code with different combinations of **maxiter**= 10, 100 and **mutprob**= 0.1, 0.5, 0.9. Observe the initial population and final population. Conclusions?

```{r,echo=FALSE}
maxiter = c(10,100)
mutprob = c(0.1,0.5,0.9)
names = c("maxiter = 10 & mutprob = 0.1",
          "maxiter = 100 & mutprob = 0.5",
          "maxiter = 10 & mutprob = 0.9",
          "maxiter = 100 & mutprob = 0.1",
          "maxiter = 10 & mutprob = 0.5",
          "maxiter = 100 & mutprob = 0.9")
pairs = data.frame(maxiter=rep(maxiter,3),mutprob=rep(mutprob,2),name=names)
pairs = split(pairs,pairs[,3])

plot(arrangeGrob(grobs=lapply(t(pairs), func4)))
```


\newpage
# Question 2: EM algorithm

```{r, echo=FALSE}
# clean the environment
rm(list=ls())
# load the data
data2 = read.csv("physical1.csv")
```

The data file *physical.csv* describes a behavior of two related physical processes $Y = Y (X) and Z = Z(X)$.


## 1. Make a time series plot 
describing dependence of $Z$ and $Y$ versus $X$. Does it seem that two processes are related to each other? What can you say about the variation of the response values with respect to $X$?

```{r, echo=FALSE}
# Z & Y versus X
col = c("Y" = "#FFC312", "Z" = "#0652DD")

data21 = reshape2::melt(data2, id.vars="X")
ggplot(data = data21, aes(x = X,y=value,color=variable)) +
  geom_line() +
  ggtitle("Z & Y versus X") + 
  ylab("Z & Y")+
  scale_color_discrete("")

```

```{r}
ggplot(data = data.frame(X=data2$X,variation=data2$Y-data2$Z), aes(x = X)) +
  geom_line(aes(y = variation)) + 
  ggtitle("variation versus X")
```

It does not seem the two processes are related to each other.
In the beginning, the Z value does sleep off more than double of Y. 
Also, the movements of the processes rarely lie on top of each other. 
We can also recognize the Z values are incomplete in some parts, the curve has gaps. 

- variation of the response values with respect to X?


## 2. Note that there are some missing values of Z in the data 
which implies problems in estimating models by maximum likelihood. Use the following model

$$ Y_i \sim \exp(X_i/\lambda),~~ \ Z_i \sim \exp(X_i/2\lambda)$$

where $\lambda$ is some unknown parameter.
*The goal is to derive an EM algorithm that estimates $\lambda$.*

We know the probability density functions (PDF) of $Z$ are
$$f(Z_i)=\frac{X_i}{2\lambda}e^{-\frac{X_i}{2\lambda}Z_i}.$$
Since we consider that all the $X_i$ are independent and we can thereby get the PDF their joint distributions, which is
$$L(Z)=\frac{\prod X_i}{2\lambda}e^{-\frac{\sum X_i}{2\lambda}\sum Z_i}.$$
Thus, their log-likelihoods should be 
$$ \log(Z|X,\lambda)=\frac{\sum \ln X_i}{\ln(2\lambda)}-\frac{\sum X_i\sum Z_i}{2\lambda}.$$
By partial derivativing with $\lambda$, we get
$$
\begin{aligned}
\nabla\lambda =& 2\cdot \frac{1}{2\lambda}\frac{\sum \ln X_i}{(\ln(2\lambda))^2}-(-2)\cdot \frac{-\sum X_i\sum Z_i}{(2\lambda)^2}\\\\
=&\frac{\sum \ln X_i}{\lambda\ln^2(2\lambda)}-\frac{\sum X_i\sum Z_i}{2\lambda^2}~~.
\end{aligned}
$$

# 3.Implement this algorithm in R
 use $\lambda_0=100$ and convergence criterion "stop if the change in $\lambda$ is less than 0:001". What is the optimal $\lambda$ and how many iterations were required to compute it?

```{r}
X <- data2$X
loglik <- function(Y, lda){sum(log(X))/(log(lda))-sum(X)*sum(Y)/(lda)}

EM <- function(Y, eps, kmax){
  # browser()
  
    # initial state
    k <- 1
    ldak <- 100
    llvalprev<-loglik(Y, ldak)
    llvalcurr<-llvalprev+10+100*eps  # make some difference
    print(c(ldak, llvalcurr))

    while ((abs(llvalprev-llvalcurr)>eps) && (k<(kmax+1))){
    	llvalprev <- llvalcurr

    	gradient <- -sum(log(X))/(ldak*(log(ldak)^2))+sum(X)*sum(Y)/(ldak^2)
    	ldak <- ldak-gradient
    
    	## Compute log-likelihood
    	llvalcurr<-loglik(Y,ldak)
    	k<-k+1
    
      print(c(ldak, llvalcurr))
    }
}

EM(Y, 0.001, 1000)
```



```{r}

```



\newpage

# Appendix
```{r ref.label=knitr::all_labels(), echo = T, eval = F}
```
